---
title: "Full Model"
author: "Live Laugh Life Expectancy"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Model Reduction

After cleaning the data, we are left with 15 explanatory variables for life expectancy, being Infant deaths, Under-five deaths, Adult mortality, Alcohol consumption, Hepatitis B, Measles, BMI, Polio, Diphtheria, Incidents of HIV, GDP per capita, Population (in millions), Thinness (ages 10–19 years), Thinness (ages 5–9 years), and Schooling.

```{r}
# load data
data <- read.csv("Life-Expectancy-Data-Updated.csv")

# fit initial model
model <- lm(Life_expectancy ~ 
              Infant_deaths +
              Under_five_deaths +
              Adult_mortality +
              Alcohol_consumption +
              Hepatitis_B +
              Measles +
              BMI +
              Polio +
              Diphtheria +
              Incidents_HIV +
              GDP_per_capita +
              Population_mln +
              Thinness_ten_nineteen_years +
              Thinness_five_nine_years +
              Schooling,
            data = data
)

summary(model)

```

After fitting the full model, we have 9 highly statistically significant variables (Infant_deaths, Under_five_deaths, Adult_mortality, Alcohol_consumption, Hepatitis_B, BMI, Incidents_HIV, GDP_per_capita, and Schooling). We have an overall p-value or 2.2e-16 Additionally, our R-squared is 0.9789 and Adjusted R-squared is 0.9788. This indicates that there might be overfitting issues within our model Next, we look at the variance inflation factors of each regressor variable to determine if there are any multicollinearity issues.

```{r}
library(car)
vif(model)
```

Multicollinearity issues, that is variables with a vif greater than 10 arise within the regressors Infant_deaths (44.329144), Under_five_deaths (44.469991), Polio (11.990081), Diphtheria (12.940367). Additionally, Thinness_ten_nineteen_years (8.949993) and Thinness_five_nine_years (8.936053) also have rather high vif's, but still bellow 10.

To combat this we will begin by removing Under_five_deaths.
```{r}
model1 <- lm(Life_expectancy ~ 
              Infant_deaths +
              Adult_mortality +
              Alcohol_consumption +
              Hepatitis_B +
              Measles +
              BMI +
              Polio +
              Diphtheria +
              Incidents_HIV +
              GDP_per_capita +
              Population_mln +
              Thinness_ten_nineteen_years +
              Thinness_five_nine_years +
              Schooling,
            data = data
)

summary(model1)

vif(model1)
```

After removing Under_five_deaths, we have the same highly statistically predictors, as well as Thinness_ten_nineteen_years being a somewhat significant predictor. Furthermore, The vif for Infant_deaths have improved. We still have a multicollinearity issue with Polio and Diphtheria, so it is in our best interest to remove one of them. We will remove polio because it has the highest p-value (0.8059) while having a large vif.
```{r}

model2 <- lm(Life_expectancy ~ 
              Infant_deaths +
              Adult_mortality +
              Alcohol_consumption +
              Hepatitis_B +
              Measles +
              BMI +
              Diphtheria +
              Incidents_HIV +
              GDP_per_capita +
              Population_mln +
              Thinness_ten_nineteen_years +
              Thinness_five_nine_years +
              Schooling,
            data = data
)

summary(model2)

vif(model2)

```

Another solution to this could possibly combining all of the vaccine percentage variables, Hepatitis_B, Measles, Polio, and Diphtheria, into one variable, Vaccine_percentage


```{r}
data$Vaccine_percentage <- rowMeans(data[, c("Hepatitis_B", "Measles", "Polio", "Diphtheria")], na.rm = TRUE)

model3 <- lm(Life_expectancy ~ 
              Infant_deaths +
              Adult_mortality +
              Alcohol_consumption +
              BMI +
              Incidents_HIV +
              GDP_per_capita +
              Population_mln +
              Thinness_ten_nineteen_years +
              Thinness_five_nine_years +
              Schooling +
              Vaccine_percentage,
            data = data
)

summary(model3)

vif(model3)

```

There does not seem to be a significant difference between these two models. If we chose to go with removing polio, we would have to remove each of the vaccines other than Diphteria as it is the only one that was statistically significant. This is likely because the vaccination somewhat consistent on each vaccine, if they had one they are likely to have the others.


Moving forward with removing Polio, our next highest p-value is measles, so we will remove that.

```{r}

model4 <- lm(Life_expectancy ~ 
              Infant_deaths +
              Adult_mortality +
              Alcohol_consumption +
              Hepatitis_B +
              BMI +
              Diphtheria +
              Incidents_HIV +
              GDP_per_capita +
              Population_mln +
              Thinness_ten_nineteen_years +
              Thinness_five_nine_years +
              Schooling,
            data = data
)

summary(model4)

vif(model4)

```
Running the new model, we can see both Diphtheria and Hepatitis_B are highly significant predictors so it is preferable to use this model rather than the average, Vaccine_percentage. The next highest p-value is Population_mlm, so we will move forward with removing that. This makes sense to not be a significatn predictor of life expectancy because there are plenty of high quality of life areas with low populations, low quality of life areas with high populations, and everything in between.

```{r}

model5 <- lm(Life_expectancy ~ 
              Infant_deaths +
              Adult_mortality +
              Alcohol_consumption +
              Hepatitis_B +
              BMI +
              Diphtheria +
              Incidents_HIV +
              GDP_per_capita +
              Thinness_ten_nineteen_years +
              Thinness_five_nine_years +
              Schooling,
            data = data
)

summary(model5)

vif(model5)

```

Thinness_five_nine_years has the next highest p-value as well as rather high vif along with Thinness_ten_nineteen_years, so it makes sense to remove Thinness_five_nine_years.

```{r}
model6 <- lm(Life_expectancy ~ 
              Infant_deaths +
              Adult_mortality +
              Alcohol_consumption +
              Hepatitis_B +
              BMI +
              Diphtheria +
              Incidents_HIV +
              GDP_per_capita +
              Thinness_ten_nineteen_years +
              Schooling,
            data = data
)

summary(model6)

vif(model6)


```
After this, all of the predictors are highly statistically significant. We still have a very high R-squared value of 0.9775. To combat this I think it may be a good idea to only have either Infant_deaths or adult mortality, as they both measure their respective deaths per 1000 in the population.

```{r}
model7 <- lm(Life_expectancy ~ 
              Adult_mortality +
              Alcohol_consumption +
              Hepatitis_B +
              BMI +
              Diphtheria +
              Incidents_HIV +
              GDP_per_capita +
              Thinness_ten_nineteen_years +
              Schooling,
            data = data
)

summary(model7)

vif(model7)

```
After removing Infant_deaths, we still have a high R-squared of 0.9591. Now we will try removing Adult_mortality and keeping Infant_deaths.

```{r}
model8 <- lm(Life_expectancy ~ 
              Infant_deaths +
              Alcohol_consumption +
              Hepatitis_B +
              BMI +
              Diphtheria +
              Incidents_HIV +
              GDP_per_capita +
              Thinness_ten_nineteen_years +
              Schooling,
            data = data
)

summary(model8)

vif(model8)

```

Once again, the R-Squared is still very high, with a value of 0.9263. I think the effort if trying to figure this out is not worth it, we will keep both variables and use model6. We opted for this model as it strikes a balance between high explanatory power and acceptable multicollinearity

# Residual Analysis

```{r}

library(MASS)

# residuals vs fitted values
plot(fitted(model6), studres(model6),
     main="Residuals vs Fitted",
     xlab="Fitted Values", ylab="Residuals")
abline(h = 0, col = "red")

# QQ plot
qqPlot(model6)

# histogram of residuals
hist(studres(model6), breaks = 15, freq = F, col="cornflowerblue",
     xlab = "Studentized Residuals")
```

The QQ plot values are very close to the line which is good. The histogram of the residuals also appears to follow a normal distribution. However, the Residuals vs. Fitted value plot indicates a transformation is needed.

# Assessment of model fit

Before we perform any transformations, we want to ensure that we have the best possible model fit. We can do this by using the ols_step_all_possible, ols_step_forward_p, and ols_step_backward_p to see if those results match up with ours. Truth be told, we did not realize those functions existed before we began reducing the model manually.

```{r}
# find all possible permutations of the full model (this is commented out because it took so long, it is saved as a csv file now)
# MS_ALL = ols_step_all_possible(model)
# write.csv(MS_ALL$result, "MS_ALL_results.csv", row.names = FALSE)
MS_ALL <- read.csv("MS_ALL_results.csv")

library(dplyr)
library(olsrr)

# find model with highest adjusted r squared
# adjr <- MS_ALL$result$adjr
# max_adjr <- which.max(adjr)
# best_model_adjr <- MS_ALL$result[max_adjr, ]
#  (commenting out original code and using new working with loaded data frame)

adjr <- MS_ALL$adjr
max_adjr <- which.max(adjr)
best_model_adjr <- MS_ALL[max_adjr, ]

best_model_adjr$predictors

# find model with lowest AIC
# aic <- MS_ALL$result$aic
# min_aic <- which.min(aic)
# best_model_aic <- MS_ALL$result[min_aic, ]

aic <- MS_ALL$aic
min_aic <- which.min(aic)
best_model_aic <- MS_ALL[min_aic, ]

best_model_aic$predictors

# find model with lowest BIC
# bic <- MS_ALL$result$sbic
# min_bic <- which.min(bic)

bic <- MS_ALL$sbic
min_bic <- which.min(bic)
min_bic == min_aic
# note: did not run best model because the best bic has the same model as the best aic

# stepwise selection with p-values
ols_stepwise_forward <- ols_step_forward_p(model)
ols_stepwise_forward$metrics$variable

ols_stepwise_backward <- ols_step_backward_p(model)
ols_stepwise_backward$metrics$variable

# fit the best models that we found
adjr_model <- lm(Life_expectancy ~ Infant_deaths+Under_five_deaths+Adult_mortality+Alcohol_consumption+Hepatitis_B+BMI+Diphtheria+Incidents_HIV+GDP_per_capita+Thinness_ten_nineteen_years+Schooling, data = data)
summary(adjr_model)
vif(adjr_model)

adjr_model1 <- lm(Life_expectancy ~ Infant_deaths+Adult_mortality+Alcohol_consumption+Hepatitis_B+BMI+Diphtheria+Incidents_HIV+GDP_per_capita+Thinness_ten_nineteen_years+Schooling, data = data)
summary(adjr_model1)

aic_bic_model <- lm(Life_expectancy ~ Infant_deaths+Under_five_deaths+Adult_mortality+Alcohol_consumption+Hepatitis_B+BMI+Incidents_HIV+GDP_per_capita+Thinness_ten_nineteen_years+Schooling, data = data)
summary(aic_bic_model)
vif(aic_bic_model)

aic_bic_model1 <- lm(Life_expectancy ~ Infant_deaths+Adult_mortality+Alcohol_consumption+Hepatitis_B+BMI+Incidents_HIV+GDP_per_capita+Thinness_ten_nineteen_years+Schooling, data = data)
summary(aic_bic_model1)
vif(aic_bic_model1)

backward_model <- lm(Life_expectancy ~ Thinness_five_nine_years+Polio+Measles+Population_mln, data = data)
summary(backward_model)
```

The models that the functions from the olsrr library provided were quite similar our model6. It is important to note that these functions do not check for multicollinearity issues. 

The model based on the highest adjusted R-squared value had predictors Infant_deaths, Under_five_deaths, Adult_mortality, Alcohol_consumption, Hepatitis_B, BMI, Diphtheria, Incidents_HIV, GDP_per_capita, Thinness_ten_nineteen_years, and Schooling. When looking at the vif's we can see Infant_deaths and Under_five_deaths are highly correlated, so it makes sense to remove Under_five_deaths as it has the higher vif. After doing so we can see that adjr_model1 is the exact same model as our main model, model6.


The model based on the lowest AIC and BIC values (remember they are the same model) had predictors Infant_deaths, Under_five_deaths, Adult_mortality, Alcohol_consumption, Hepatitis_B, BMI, Incidents_HIV, GDP_per_capita, Thinness_ten_nineteen_years, and Schooling. Notice that this contains the same exact values as adj_model other than the fact that it omits Diphtheria. After removing Under_five_deaths once more due to multicollinearity, we are left with the final model based on the AIC/BIC values, aic_bic_model1. Simillarly to our other models, this model's adjusted R-squared was 0.9773

The model obtained by the forward stepwise method had the same predictors as the adjr_model, meaning after cleaning it will be the same model as our main model, model6.

The model obtained by the backward stepwise method was the most different out of every model and had predictors Thinness_five_nine_years, Polio, Measles, and Population_mln. Interestingly enough, these are all variables that were removed from other models. This was the worst fit model, with an adjusted R-squared of 0.5167.

Taking all of this into consideration, the best choice of action is likely to go with aic_bic_model1, as the change caused by Diphtheria was rather insignificant.


# Transformations

```{r}

# fit model taking log(Life_expectancy)

log_aic_bic_model1 <- lm(log(Life_expectancy) ~ 
              Infant_deaths +
              Adult_mortality +
              Alcohol_consumption +
              Hepatitis_B +
              BMI +
              Incidents_HIV +
              GDP_per_capita +
              Thinness_ten_nineteen_years +
              Schooling,
            data = data
)

summary(log_aic_bic_model1)

```

# WAIT 

By including the predictors that relate directly to death we have essentially just found that when less adults/babies die, the life expectancy is longer. WOW!!! This is not exactly the information we wanted to capture, it is a pretty obvious assumption, so it makes sense to try to refit the model with out it. So yes, everything up til now was all in vain. 

# New Initial Model Selection

```{r}
new_model <- lm(Life_expectancy ~ 
              Alcohol_consumption +
              Hepatitis_B +
              Measles +
              BMI +
              Polio +
              Diphtheria +
              Incidents_HIV +
              GDP_per_capita +
              Population_mln +
              Thinness_ten_nineteen_years +
              Thinness_five_nine_years +
              Schooling +
              Economy_status_Developed,
            data = data
)

summary(new_model)

vif(new_model)

# find all possible permutations of the new model 
# new_MS_ALL = ols_step_all_possible(new_model)
# write.csv(new_MS_ALL$result, "new_MS_ALL_results.csv", row.names = FALSE)
new_MS_ALL <- read.csv("new_MS_ALL_results.csv")

new_adjr <- new_MS_ALL$adjr
new_max_adjr <- which.max(new_adjr)
new_best_model_adjr <- new_MS_ALL[new_max_adjr, ]
new_best_model_adjr$predictors

new_aic <- new_MS_ALL$aic
new_min_aic <- which.min(new_aic)
new_best_model_aic <- new_MS_ALL[new_min_aic, ]
new_best_model_aic$predictors

new_max_adjr == new_min_aic

new_bic <- new_MS_ALL$sbic
new_min_bic <- which.min(new_bic)
new_best_model_bic <- new_MS_ALL[new_min_bic, ]
new_best_model_bic$predictors

new_min_bic == new_max_adjr

# stepwise selection with p-values
new_ols_stepwise_forward <- ols_step_forward_p(new_model)
new_ols_stepwise_forward$metrics$variable


new_model1 <- lm(Life_expectancy ~ Alcohol_consumption+Hepatitis_B+Measles+BMI+Polio+Diphtheria+Incidents_HIV+GDP_per_capita+Population_mln+Thinness_ten_nineteen_years+Thinness_five_nine_years+Schooling+Economy_status_Developed, data = data)

summary(new_model1)
```

After removing the death variables, the best model, which has the highest adjusted R-squared and lowest AIC and BIC values is the new full model.
Now we will begin reducing this model, Thinness_ten_nineteen_years is not statistically significant so we will remove it.

```{r}

new_model1 <- lm(Life_expectancy ~ 
              Alcohol_consumption +
              Hepatitis_B +
              Measles +
              BMI +
              Polio +
              Diphtheria +
              Incidents_HIV +
              GDP_per_capita +
              Population_mln ++
              Thinness_five_nine_years +
              Schooling +
              Economy_status_Developed,
            data = data
)

summary(new_model1)

vif(new_model1)

```

Now, every explanatory variable is statistically significant, however, there is a multicollinearity issue between Polio and Diphtheria, so we will test models by removing each of them and see which one has a higher adjusted R-squared.

```{r}

new_model2 <- lm(Life_expectancy ~ 
              Alcohol_consumption +
              Hepatitis_B +
              Measles +
              BMI +
              Diphtheria +
              Incidents_HIV +
              GDP_per_capita +
              Population_mln ++
              Thinness_five_nine_years +
              Schooling +
              Economy_status_Developed,
            data = data
)

summary(new_model2)

vif(new_model2)

new_model3 <- lm(Life_expectancy ~ 
              Alcohol_consumption +
              Hepatitis_B +
              Measles +
              BMI +
              Polio +
              Incidents_HIV +
              GDP_per_capita +
              Population_mln ++
              Thinness_five_nine_years +
              Schooling +
              Economy_status_Developed,
            data = data
)

summary(new_model3)

vif(new_model3)

```
When dropping Diphtheria, Hepatitis-B becomes in significant, so we will drop Polio instead, so we will go with new_model2 as our reduced model. It has an adjusted R-squared 0f 0.83 which is much more preferable to the previous 0.97, it indicated that these values have a strong relation to Life_expectancy without over fitting the model.

# New Residual Analysis

```{r}

# residuals vs fitted values
plot(fitted(new_model2), studres(new_model2),
     main="Residuals vs Fitted",
     xlab="Fitted Values", ylab="Residuals")
abline(h = 0, col = "red")

# QQ plot
qqPlot(new_model2)

# histogram of residuals
hist(studres(new_model2), breaks = 10, freq = F, col="cornflowerblue",
     xlab = "Studentized Residuals")

# cook's distance

cooks_d <- cooks.distance(new_model2)
plot(cooks_d)

# remove influential points flagged by cook's distance

n <- nrow(data)

influential_point <- which(cooks_d > (4 / n))

cleaned_data <- data[-influential_point,]

cleaned_new_model2 <- lm(Life_expectancy ~ 
              Alcohol_consumption +
              Hepatitis_B +
              Measles +
              BMI +
              Diphtheria +
              Incidents_HIV +
              GDP_per_capita +
              Population_mln ++
              Thinness_five_nine_years +
              Schooling +
              Economy_status_Developed,
            data = cleaned_data
)

summary(cleaned_new_model2)

cleaned_cooks_d <- cooks.distance(cleaned_new_model2)
plot(cleaned_cooks_d)

library(car)
library(MASS)

# residuals vs fitted values
plot(fitted(cleaned_new_model2), studres(cleaned_new_model2),
     main="Residuals vs Fitted",
     xlab="Fitted Values", ylab="Residuals")
abline(h = 0, col = "red")

# QQ plot
qqPlot(cleaned_new_model2)

# histogram of residuals
hist(studres(cleaned_new_model2), breaks = 10, freq = F, col="cornflowerblue",
     xlab = "Studentized Residuals")

```

# New Transformations

```{r}

# log transformation

log_cleaned_new_model2 <- lm(log(Life_expectancy) ~ 
              Alcohol_consumption +
              Hepatitis_B +
              Measles +
              BMI +
              Diphtheria +
              Incidents_HIV +
              GDP_per_capita +
              Population_mln ++
              Thinness_five_nine_years +
              Schooling +
              Economy_status_Developed,
            data = cleaned_data
)

summary(log_cleaned_new_model2)


# residuals vs fitted values
plot(fitted(log_cleaned_new_model2), studres(log_cleaned_new_model2),
     main="Residuals vs Fitted",
     xlab="Fitted Values", ylab="Residuals")
abline(h = 0, col = "red")

# QQ plot
qqPlot(log_cleaned_new_model2)

# histogram of residuals
hist(studres(log_cleaned_new_model2), breaks = 10, freq = F, col="cornflowerblue",
     xlab = "Studentized Residuals")

# square root transformation

sqrt_cleaned_new_model2 <- lm(sqrt(Life_expectancy) ~ 
              Alcohol_consumption +
              Hepatitis_B +
              Measles +
              BMI +
              Diphtheria +
              Incidents_HIV +
              GDP_per_capita +
              Population_mln ++
              Thinness_five_nine_years +
              Schooling +
              Economy_status_Developed,
            data = cleaned_data
)

summary(log_cleaned_new_model2)


# residuals vs fitted values
plot(fitted(sqrt_cleaned_new_model2), studres(sqrt_cleaned_new_model2),
     main="Residuals vs Fitted",
     xlab="Fitted Values", ylab="Residuals")
abline(h = 0, col = "red")

# QQ plot
qqPlot(sqrt_cleaned_new_model2)

# histogram of residuals
hist(studres(sqrt_cleaned_new_model2), breaks = 10, freq = F, col="cornflowerblue",
     xlab = "Studentized Residuals")

# box-cox transformation

boxcox_result <- boxcox(cleaned_new_model2)
lambda <- boxcox_result$x[which.max(boxcox_result$y)]

boxcox1_cleaned_new_model2 <- lm((Life_expectancy ^ lambda)  ~ 
              Alcohol_consumption +
              Hepatitis_B +
              Measles +
              BMI +
              Diphtheria +
              Incidents_HIV +
              GDP_per_capita +
              Population_mln ++
              Thinness_five_nine_years +
              Schooling +
              Economy_status_Developed,
            data = cleaned_data
)

summary(boxcox1_cleaned_new_model2)

# residuals vs fitted values
plot(fitted(boxcox1_cleaned_new_model2), studres(boxcox1_cleaned_new_model2),
     main="Residuals vs Fitted",
     xlab="Fitted Values", ylab="Residuals")
abline(h = 0, col = "red")

# QQ plot
qqPlot(boxcox1_cleaned_new_model2)

# histogram of residuals
hist(studres(boxcox1_cleaned_new_model2), breaks = 10, freq = F, col="cornflowerblue",
     xlab = "Studentized Residuals")

boxcox2_cleaned_new_model2 <- lm((((Life_expectancy ^ lambda) - 1) / 2)   ~ 
              Alcohol_consumption +
              Hepatitis_B +
              Measles +
              BMI +
              Diphtheria +
              Incidents_HIV +
              GDP_per_capita +
              Population_mln ++
              Thinness_five_nine_years +
              Schooling +
              Economy_status_Developed,
            data = cleaned_data
)

summary(boxcox2_cleaned_new_model2)

# residuals vs fitted values
plot(fitted(boxcox2_cleaned_new_model2), studres(boxcox2_cleaned_new_model2),
     main="Residuals vs Fitted",
     xlab="Fitted Values", ylab="Residuals")
abline(h = 0, col = "red")

# QQ plot
qqPlot(boxcox2_cleaned_new_model2)

# histogram of residuals
hist(studres(boxcox2_cleaned_new_model2), breaks = 10, freq = F, col="cornflowerblue",
     xlab = "Studentized Residuals")

```

Now that we have transformed variables, we will compare their AIC values to determine which is best.

```{r}

AIC(cleaned_new_model2, log_cleaned_new_model2, sqrt_cleaned_new_model2, boxcox1_cleaned_new_model2, boxcox2_cleaned_new_model2)
BIC(cleaned_new_model2, log_cleaned_new_model2, sqrt_cleaned_new_model2, boxcox1_cleaned_new_model2, boxcox2_cleaned_new_model2)

```

The log transformation has the lowest AIC and BIC values, indicating that it is the best model out of the original and these transformations. The tails of the QQ plot are not as good as the original mode but we think this is a valid trade off for the better AIC/BIC values.

```{r}



```